---
title: "Get started with tabularray"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with tabularray}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = ""
)
```

```{r setup, message = FALSE}
library(tabularray)
library(dplyr)

# helper to display the generated LaTeX in this vignette
show_latex <- function(x) cat(as.character(tblr_as_latex(x)))
```

`tabularray` turns a data frame into LaTeX markup for the
[tabularray](https://ctan.org/pkg/tabularray) LaTeX package. Inspired by
[gt](https://gt.rstudio.com), you build a table by piping a data frame into
`tblr()` and then chaining `set_*()` functions to add styling, before rendering
with `tblr_as_latex()`.

## A first table

`tblr()` initializes a data frame as a table object. Printing it (or calling
`tblr_as_latex()`) produces LaTeX:

```{r}
df <- tibble(
  continent = c("Europe", "Asia", "Asia"),
  country   = c("Germany", "China", "Afghanistan"),
  value     = c(0.17, 0.23, 11.3)
)

df |> tblr() |> show_latex()
```

Numeric columns are right-aligned, text columns left-aligned. Column labels
default to the column names.

## Numbers

`tabularray` does not try to be a number-formatting toolkit. For currency,
percentages, thousands separators, significant digits, or any other per-column
formatting, format the column before calling `tblr()` — for example with
[scales](https://scales.r-lib.org), `base::formatC()`, or `base::sprintf()`:

```{r}
tibble(
  item  = c("Widget", "Gadget"),
  price = c(1999.5, 12.4)
) |>
  mutate(price = sprintf("$%.2f", price)) |>  # format first ...
  tblr() |>                                   # ... then build the table
  show_latex()
```

The formatted `price` is now a text column, so the literal `$` it contains is
escaped to `\$` for you — write the plain character you want and let `tblr`
handle the escaping.

Columns you leave as numbers get a sensible default: two decimal places, with
integer and whole-valued columns printed without decimals.

```{r}
tibble(
  year  = c(2021L, 2022L, 2023L),
  share = c(3.04, 12.5, 100.5)
) |>
  tblr() |>
  show_latex()
```

## Escaping

Characters that are special in LaTeX (`&`, `%`, `$`, `_`, `#`, `{`, `}`, `~`,
`^`, `\`) are escaped automatically in **text columns**, so ordinary data is
safe:

```{r}
tibble(account = c("R&D", "50% off", "a_b")) |>
  tblr() |>
  show_latex()
```

Two things to keep in mind:

* **Convert column types before `tblr()`.** Escaping is decided from each
  column's type when the table is rendered. If you format a column upstream (as
  in the price example above), it is treated as text and escaped like any other
  text column.

* **Labels, spanners, alignment, notes, and options are raw LaTeX.** Values you
  pass to `set_column_labels()`, `set_column_spanner()`, `set_alignment()`,
  `set_source_notes()`, `set_interface()`, and `set_options()` are inserted
  verbatim, so you can use LaTeX commands there — and you must escape any special
  characters yourself (e.g. write `R\&D`, not `R&D`).

```{r}
df |>
  tblr() |>
  set_column_labels(value = "\\textbf{Value}") |>
  show_latex()
```

## Grouping, spanners, and themes

Group rows with `dplyr::group_by()` before `tblr()`, add spanning headers with
`set_column_spanner()` (call it once per spanner row, bottom-up), and tune the
appearance with `set_theme()`:

```{r}
starwars |>
  filter(homeworld == "Tatooine", !name %in% c("Anakin Skywalker", "Darth Vader")) |>
  select(name, height, mass, sex, birth_year) |>
  mutate(sex = stringr::str_to_title(sex)) |>
  group_by(sex) |>
  tblr(type = "float", caption = "Tatooine characters") |>
  set_column_labels(name = "", birth_year = "Birth Year") |>
  set_column_spanner(c(height, mass) ~ "Physical") |>
  set_theme(row_group_style = "panel") |>
  show_latex()
```

The `type` argument selects the LaTeX environment: `"simple"` (default, inline),
`"float"` (numbered, captioned, supports notes), or `"break"` (page-breaking).

## Using the output

The generated markup needs the `tabularray` package and its `booktabs` library.
In a LaTeX preamble:

```latex
\usepackage{tabularray}
\UseTblrLibrary{booktabs}
```

A reasonably recent `tabularray` (2023 or later) is recommended. In Quarto or R
Markdown, the output is embedded verbatim — see the
[README](https://github.com/turbanisch/tabularray) for the YAML setup. You can
also copy the LaTeX into any editor.
